---
title: "Reproducible Research: Peer Assessment 1"
---
by Jayakumar Nagarathinam

The following sections explains the loading, preprocessing and processing of data.

## Loading and preprocessing the data

```{r}
unzip(zipfile = "activity.zip")
data <- read.csv("activity.csv",header = T)

```

## What is mean total number of steps taken per day?

Code and Output:

```{r}
library(ggplot2)
totalSteps <- tapply(data$steps,data$date, FUN = sum, na.rm = TRUE)
qplot(totalSteps, binwidth=1000, xlab="total number of steps taken each day")
mean(totalSteps, na.rm=TRUE)
median(totalSteps, na.rm=TRUE)
```

## What is the average daily activity pattern?

```{r}
library(ggplot2)
dailyAverage <- aggregate(x=list(steps=data$steps), 
                          by=list(interval=data$interval), 
                          FUN=mean, na.rm=TRUE)
ggplot(data=dailyAverage, aes(x=interval, y=steps)) +
       geom_line()                              +
       xlab("5-minute interval")                +
       ylab("Average Number of Steps Taken")
```

On average, does the 5 minute interval contains the maximum number of steps?

```{r}
dailyAverage[which.max(dailyAverage$steps),]
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
missing <- is.na(data$steps)
# How many missing
table(missing)
```

All of the missing values are filled in with mean value for that 5-minute interval.
```{r}
# Replace each missing value with the mean value of its 5-minute interval
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (dailyAverage[dailyAverage$interval==interval, "steps"])
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```
Now, using the filled data set, let's make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps.

```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps)
median(total.steps)
```

## Are there differences in activity patterns between weekdays and weekends?

Firstly, let's find the day of the week for each measurement in the dataset. In this part, we use the dataset with the filled-in values.

```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```

Now, let's make a panel plot containing plots of average number of steps taken on weekdays and weekends.

```{r}
averages <- aggregate(steps ~ interval + day, data=filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```